---
title: "How Many Fast Food Restaurants per Capita Are in Each State/County?"

author: "Caitlin Uang"
date: "`r Sys.Date()`"
format:
  html:
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    toc: true
editor: visual
execute:
  warning: false
  message: false
  echo: true
---

## Introduction

## Data Acquisition
```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(tigris)
library(sf)
library(DT)
```

### NaNDA: Eating and Drinking Places by Census Tract & ZCTA, 1990–2021

To acquire the national fast food count per state and county, this project will use the [NaNDA: Eating and Drinking Places by Census Tract & ZCTA, 1990-2021](https://www.icpsr.umich.edu/sites/nanda/view/studies/208751) dataset and filter year 2019. The dataset at census tract level is used. NaNDA is the National Neighborhood Data Archive, providing counts and density of food establishments (restaurants, bars, fast food) from 1990-2021, sourced from the National Establishment Time Series (NETS) database.

```{r}
# Cache shapefiles locally (faster repeat runs)
options(tigris_use_cache = TRUE)

# Load dataset and use only 2019 fast food count
eatdrink_tract <- read_csv("nanda_eatdrink_Tract20_1990-2021_01P.csv")

fastfood_raw <- eatdrink_tract |>
  filter(year == 2019) |>
  arrange(desc(count_fastfood)) |>
  select(tract_fips20,totpop,count_fastfood)

```

### U.S. Census Bureau ACS B01003 (2019)

[The U.S. Census Bureau American Community Survey (ACS) Table B01003](https://data.census.gov/table/ACSDT5Y2019.B01003?q=Population+Total&g=010XX00US$1400000) provides the total population for a given geographic area. The ACS B01003 table comes from 2019. 

```{r}
# Load dataset
pop_2019 <- read_csv("ACSDT5Y2019.B01003-Data.csv")
```

### Join Both Datasets

The NaNDA dataset uses tract FIPS (Federal Information Processing Standards) to identify each census tract within the U.S., while ACS B01003 uses GEOID, a full numeric code, to uniquely identify all administrative and statistical geographic areas. GEOIDs encompasses FIPS codes and will be used to join both datasets. 
```{r}

# Get FIPS from GEOID
pop_clean <- pop_2019 |>
  mutate(tract_fips20 = substr(GEO_ID, nchar(GEO_ID) - 10, nchar(GEO_ID)))

# Join ACS and fast food dataset by FIPS code and filter out tract with 0 population
fastfood_tract <- fastfood_raw |>
  left_join(pop_clean, by = "tract_fips20") |>
  select(tract_fips20,GEO_ID,totpop,B01003_001E,count_fastfood) |>
  rename(acs_pop = B01003_001E) |>
  mutate(acs_pop = as.numeric(acs_pop)) |>
  filter(acs_pop > 0)
```

## Data Exploration

Let's look at the joined dataset. Now that we have both dataset combined, we can calculate the 

```{r}
fastfood_tract |>
  select(
    FIPS = tract_fips20,
    GEOID = GEO_ID,
    `Total Population` = acs_pop,
    `Fast Food Count` = count_fastfood
  ) |>
  datatable(
    caption = "Fast Food Restaurants and Population by Census Tract",
    rownames = FALSE,
    options = list(
      pageLength = 10,
      scrollX = TRUE
    )
  ) |>
  formatRound("Total Population", 0, mark = ",") |>
  formatRound("Fast Food Count", 0)
          
```

### Summary Statistics by Census Tract

#### Fast Food Count

The distribution of fast food restaurants across census tract is highly right skewered as there is a large number of census tracts with no fast food restaurants. As the number of restaurants increase, the number of census tract decreases rapidly. 
```{r}
summary(fastfood_tract$count_fastfood)

fastfood_tract |>
  ggplot(aes(x = count_fastfood)) +
  geom_histogram(
    bins = 50,
    fill = "#9badff",
    color = "white",
    linewidth = 0.3
  ) +
  labs(
    title = "Distribution of Fast Food Count by Census Tract",
    caption = "Source: NaNDA Eating and Drinking Places by Census Tract & ZCTA, 1990–2021\nU.S. Census Bureau ACS B01003 (2019)",
    x = "Fast Food Count",
    y = "# of Census Tracts"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0)
  )

```

#### Total Population
The distribution of population across census tracts is relatively contained and mildly right skewed between 2,000 and 6,000 people and very few extreme outliers. The median falls at 3,852, representing the population of a typical census tract.
```{r}
summary(fastfood_tract$acs_pop)

fastfood_tract |>
  ggplot(aes(x = acs_pop)) +
  geom_histogram(
    bins = 50,
    fill = "#9badff",
    color = "white",
    linewidth = 0.3
  ) +
  geom_vline(
    xintercept = median(fastfood_tract$acs_pop, na.rm = TRUE),
    linetype = "dashed",
    linewidth = 0.8,
    color = "#0432FF"
  ) +
  labs(
    title = "Distribution of Population by Census Tract",
    caption = "Source: NaNDA Eating and Drinking Places by Census Tract & ZCTA, 1990–2021\nU.S. Census Bureau ACS B01003 (2019)",
    x = "Population",
    y = "# of Census Tracts"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0)
  )

```
Now that we have a better understanding of the distribution of fast food counts and population by census tract, the team shifted focus on analyzing the fast food density, measured by count of fast food restaurants per 1,000 people to have a relative comparison of fast food access across census tracts. During the early days of this project, the team found that looking only at just the count of fast food restaurants is misleading because census tracts vary in the number of people living in them, while fast food restaurants are unevenly distributed. For example, states with larger total population have larger amount of fast food restaurants available. Using fast food density allows for a more meaning comparison as it accounts for population size.

In the next section, we will look at fast food density at the county and state level to examine any patterns that can help us answer the overarching question.

## Fast Food Density by State

To examine at the national scale, we first aggregated fast food restaurant counts from the census tract level to the state level. Looking at the data at a state level reduces local variability caused by small population census tracts and also provide a clearer view of regional differences in fast food restaurant availability. For a fa



```{r}
# Aggregate tract to state
fastfood_state <- fastfood_tract |>
  mutate(state_fips20 = substr(tract_fips20, 1, 2)) |>
  group_by(state_fips20) |>
  summarize(
    total_fastfood = sum(count_fastfood, na.rm = TRUE),
    total_pop      = sum(acs_pop, na.rm = TRUE),
    den_fastfood   = (total_fastfood / total_pop) * 1000,
    .groups = "drop"
  ) |>
  filter(total_pop > 0)

# Load state shapefile and join
state_shapes <- states(year = 2019, cb = TRUE) |>
  mutate(STATEFP = as.character(STATEFP))

fastfood_map_state <- state_shapes |>
  left_join(fastfood_state, by = c("STATEFP" = "state_fips20"))
```
#### Top 10 States with the Highest Fast Food Density
```{r}
# Create datatable
datatable(
  fastfood_map_state |>
    st_drop_geometry() |>
    arrange(desc(den_fastfood)) |>
    slice_max(den_fastfood, n = 10) |>
    select(
      State = NAME,
      `Fast Food Density` = den_fastfood,
      `Total Population` = total_pop, 
      `Fast Food Count` = total_fastfood),
  rownames = FALSE,
  caption = "Top 10 States with the Highest Fast Food Density",
  options = list(
    searching = FALSE,
    scrollX = FALSE,
    autoWidth = TRUE
  )
) |>
  formatRound("Fast Food Density", 2) |>
  formatRound("Total Population", 0, mark = ",") |>
  formatRound("Fast Food Count", 0)
```

#### Heat Map ofFast Food Density by State Map

```{r}
# Create heat map 
ggplot(fastfood_map_state) +
  geom_sf(aes(fill = den_fastfood), color = "white") +
  scale_fill_gradient(
    low = "#ffffff",
    high = "#932092",
    name = "Fast Food\nDensity", limits = c(NA,NA)
                       )+
  coord_sf(
    xlim = c(-125, -66),
    ylim = c(24, 50),
    expand = FALSE
  ) +
  labs(
    title = "Fast Food Restaurants per 1,000 People",
    subtitle = "State, 2019",
    caption = "Source: NaNDA Eating and Drinking Places by Census Tract & ZCTA, 1990–2021\nU.S. Census Bureau ACS B01003 (2019)"
  ) +
  
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    plot.margin = margin(5, 5, 5, 5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0)
  )
```

#### Top 10 States Highest Fast Food Concentration

```{r}
# Rank and flag top 10 states
fastfood_map_state_topflag <- fastfood_map_state |>
  arrange(desc(den_fastfood)) |>
  mutate(
    rank = row_number(),
    den_top10 = ifelse(rank <= 10, den_fastfood, NA_real_)
  )

ggplot(fastfood_map_state_topflag) +
  geom_sf(aes(fill = den_top10),
          color = "white",
          linewidth = 0.3) +
  scale_fill_distiller(
    palette = "Blues",      # Nice clean blues
    direction = 1,          # Light → dark
    na.value = "#e0e0e0",   # Light grey for non–top 10
    name = "Fast-Food per\n1,000 People"
  ) +
  labs(
    title = "Top 10 States with the Most Fast-Food Restaurants per 1,000 People",
    subtitle = "Non–Top 10 States Shown in Grey",
    caption = "Source: NaNDA Tract20 (1990–2021) & U.S. Census ACS 2019"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 9, hjust = 0),
    plot.margin = margin(10, 10, 10, 10)
  )

```






### Fast Food Restaurants per 1,000 people by County


```{r}
#State look up
state_lookup <- states(year = 2019) |>
  st_drop_geometry() |>
  select(STATEFP, STUSPS, NAME) |>
  rename(
    State_Abbr = STUSPS,
    State_Name = NAME
  )

# Aggregate tract to county
den_fastfood_county <- fastfood_tract |>
  mutate(county_fips20 = substr(tract_fips20, 1, 5)) |>
  group_by(county_fips20) |>
  summarize(
    total_fastfood = sum(count_fastfood, na.rm = TRUE),
    total_pop      = sum(acs_pop, na.rm = TRUE),
    den_fastfood   = (total_fastfood / total_pop) * 1000,
    .groups = "drop"
  ) |>
  filter(total_pop > 0)

# Load shapefile and join

county_shapes <- counties(year = 2019, cb = TRUE) |>
  mutate(GEOID = as.character(GEOID))

fastfood_map_county <- county_shapes |>
  left_join(den_fastfood_county, by = c("GEOID" = "county_fips20")) |>
  left_join(state_lookup, by = "STATEFP") |>
  mutate(
    den_fastfood_plot = ifelse(den_fastfood == 0, NA, den_fastfood)
  )

# Create datatable
datatable(
  fastfood_map_county |>
    st_drop_geometry() |>
    arrange(desc(den_fastfood)) |>
    select(
      County = NAME,
      State = State_Name,
      `Fast Food Count` = total_fastfood,
      `Total Population` = total_pop,
      `Fast Food Density` = den_fastfood
    ),
  caption = "Fast Food Restaurants per 1,000 People by County (2019)",
  rownames = FALSE,
  options = list(
    pageLength = 10,
    scrollX = TRUE
  )
) |>
  formatRound("Fast Food Count", 0, mark = ",") |>
  formatRound("Total Population", 0, mark = ",") |>
  formatRound("Fast Food Density", 2)
```
#### Distribution of Fast Food Density per County
```{r}
fastfood_map_county |>
  st_drop_geometry() |>
  ggplot(aes(x = den_fastfood)) +
  geom_histogram(
    bins = 30,
    fill = "#9badff",
    color = "white",
    linewidth = 0.3
  ) +
  coord_cartesian(xlim = c(0, 1)) + 
  labs(
    title = "Fast Food Restaurants per 1,000 People",
    subtitle = "Distribution of Fast Food Density per County, 2019",
    caption = "Source: NaNDA Eating and Drinking Places by Census Tract & ZCTA, 1990–2021\nU.S. Census Bureau ACS B01003 (2019)",
    x = "Fast Food Density",
    y = "# of Counties"
  ) +
  theme(
    axis.line = element_line(color = "black", linewidth = 0.4),
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    plot.margin = margin(5, 5, 5, 5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0),
      panel.background = element_rect(fill = "white")

  )


```
#### County Map
```{r}
ggplot(fastfood_map_county) +
  geom_sf(aes(fill = den_fastfood), color = NA, linewidth = 0.1) +
  
  scale_fill_gradientn(
  colours = c("grey90", "#e792e7", "#5b7cff", "#0432FF"),
  name = "Fast Food\nDensity",
  limits = c(NA, NA),
  na.value = "#ffffff",
  trans = "sqrt"
  ) + 

  coord_sf(
    xlim = c(-125, -66),
    ylim = c(24, 50),
    expand = FALSE
  ) +
  
  labs(
    title = "Fast Food Restaurants per 1,000 People",
    subtitle = "County, 2019",
    caption = "Source: NaNDA Eating and Drinking Places by Census Tract & ZCTA, 1990–2021\nU.S. Census Bureau ACS B01003 (2019)"
  ) +
  
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    plot.margin = margin(5, 5, 5, 5),
    plot.caption = element_text(size = 10, face = "italic", hjust = 0)
  )


```




